Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-19 paper.tex
Make paragraph titles bold and upright

\newcommand{\xcom}[1]{{\textcolor{cyan}{Xavier: #1}} }
\newcommand{\jcom}[1]{{\textcolor{orange}{Jannis: #1}} }

@@

\paragraph{Forward State.}
We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
Let $\States$ be the set of forward states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The forward state's interface then consists of two functions
\begin{equation*}
  \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
\end{equation*}
The $\addHyp$ function adds a hypothesis to the forward state.
Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
The second output of $\addHyp$, $C$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

@@

\paragraph{Rule State.}
Partial matches are specific to each rule, so the forward state naturally decomposes into a set of \emph{rule states}, one per rule.
Given a goal $Γ ⊢ A$ and a rule $r$, the rule state for $r$ contains exactly the partial matches of $r$ in context $Γ$.
A rule state supports $\addHyp$ and $\delHyp$ methods analogous to those on forward states.
Adding a hypothesis to a forward state then amounts to adding it to each rule state (which may yield new complete matches for the rule); deleting a hypothesis amounts to deleting it from each rule state.
We initially focus on how to add hypotheses, since this question motivates much of the design of the rule state data structure.

@@

\paragraph{Variable Indices.}
Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
In general, to extend a match $m$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $m$, $\sub(m)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.

@@

\paragraph{Variable Clusters.}
We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
As a result, our above algorithms fail.
Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $[h₁]$.
The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $[h₁]$ can be extended with $h₂$.
In fact, any match at level~1 can be extended with any hypothesis matching slot~2 since the two slots are independent.

@@

\paragraph{Deletion.}
When a hypothesis $h$ is removed from a goal's context, we must update the forward state accordingly.
This means deleting $h$ from any rule state to which it was previously added, i.e.\ from any rule state for a rule $r$ with a slot $i$ matching $h$.
To facilitate deletion, we cache this information when $h$ is added to the rule state.
When $h$ is deleted, we can then efficiently determine the relevant rule states and delete $h$ from the hypothesis index $H$ at slot $i$ and from the match index $M$ at all slots $j ≥ i$.

\paragraph{Example 1.}
As a first example, consider a rule state for the transitivity rule $r₁ : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
$h₁ : b ≤ c$ and $h₂ : a ≤ b$.
The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.

@@

\paragraph{Example 2.}
We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot~2.
There are no matches yet, so there is nothing to extend.
We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the following rule state:
\begin{align*}
  H_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}};
  \end{tikzpicture}
  &&
  H_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}};
  \end{tikzpicture}
  &&
  M_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$a$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$b$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align*}

@@

\paragraph{Unification.}
In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
As a result, unifying premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
Our technique therefore seeks to minimise the number of unifications performed.

@@

\paragraph{RPINF.}
Since the hypothesis index $H$ and the match index $M$ of a rule state have the same domain, we combine them into one map.
This map, whose domain is a set of triples $(x, i, t)$, is represented by three nested persistent hash maps.

@@

\paragraph{Match Equivalence.}
Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.

\paragraph{Redundant Hypotheses.}
When we find a complete match, we add a new hypothesis of the corresponding type $T$ to the goal, but only if there is no other hypothesis of type $T$ already present.
To efficiently implement this check, we store in the forward match state a set containing the RPINFs of all hypotheses that have been added to (and not subsequently removed from) the state.
We modified the naive forward reasoning implementation to use the same mechanism.

\paragraph{Lazy Insertion.}
Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
We therefore delay insertions until the index has selected at least one hypothesis for each slot.
This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process hypotheses for $a$ if there is, in fact, an involutive function in the context.

@@

\paragraph{Transitivity Benchmark.}
Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$, where either $a = 0$ (yielding easy unification problems) or $a = 100$.
Saturating this goal adds $n(n-1)/2$ hypotheses to the context.

@@

\paragraph{Independence Benchmark.}
In the last example, the incremental algorithm performs well partly because each match at level~1 can be combined with many hypotheses at level~2 without processing the level~1 hypothesis repeatedly.
We now consider a benchmark with many independent rules where this kind of reuse does not occur.

@@

\paragraph{Depth Benchmark.}
Our final benchmark concerns situations where many rules match almost, but not quite.
We use 100 copies of a rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
In other words, the hypotheses match all premises with compatible substitutions, except the $k^{\text{th}}$ premise from the back.
Since the naive algorithm processes the premises back-to-front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.

@@

\paragraph{White-Box Tactics.}
Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{PVS} have special support for forward reasoning.
Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.

@@

\paragraph{SATCHMO.}
The SATCHMO first-order theorem prover~\cite{SATCHMO} uses a search procedure that treats a given set of implications as forward rules and uses a case-splitting rule to perform case analysis on disjunctive hypotheses.
SATCHMO-style provers may benefit from our incremental approach if the set of implications is large.
However, since unification is comparatively cheap in first-order logic, it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the forward state.
SATCHMO also employs an optimisation that \enquote*{focuses} forward reasoning on the hypothesis established by the last case split, providing a different form of incrementality.

\paragraph{Knowledge-Based Systems.}
Some knowledge-based reasoning systems, e.g.\ Algernon~\cite{Algernon} and EYE~\cite{EYE}, support combined forward and backward reasoning.
However, to our knowledge, no such system allows non-forward rules to change the assumptions available to forward rules, so they do not require incrementality.
Nevertheless, our technique bears some resemblance to the RETE algorithm~\cite{RETE} for forward reasoning in large knowledge bases.

>>>>

\newcommand{\para}[1]{\paragraph{\bfseries\upshape #1}}
\newcommand{\xcom}[1]{{\textcolor{cyan}{Xavier: #1}} }
\newcommand{\jcom}[1]{{\textcolor{orange}{Jannis: #1}} }

@@

\para{Forward States.}
We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
Let $\States$ be the set of forward states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The forward state's interface then consists of two functions
\begin{equation*}
  \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
\end{equation*}
The $\addHyp$ function adds a hypothesis to the forward state.
Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
The second output of $\addHyp$, $C$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

@@

\para{Rule States.}
Partial matches are specific to each rule, so the forward state naturally decomposes into a set of \emph{rule states}, one per rule.
Given a goal $Γ ⊢ A$ and a rule $r$, the rule state for $r$ contains exactly the partial matches of $r$ in context $Γ$.
A rule state supports $\addHyp$ and $\delHyp$ methods analogous to those on forward states.
Adding a hypothesis to a forward state then amounts to adding it to each rule state (which may yield new complete matches for the rule); deleting a hypothesis amounts to deleting it from each rule state.
We initially focus on how to add hypotheses, since this question motivates much of the design of the rule state data structure.

@@

\para{Variable Indices.}
Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
In general, to extend a match $m$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $m$, $\sub(m)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.

@@

\para{Variable Clusters.}
We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
As a result, our above algorithms fail.
Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $[h₁]$.
The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $[h₁]$ can be extended with $h₂$.
In fact, any match at level~1 can be extended with any hypothesis matching slot~2 since the two slots are independent.

@@

\para{Deletion.}
When a hypothesis $h$ is removed from a goal's context, we must update the forward state accordingly.
This means deleting $h$ from any rule state to which it was previously added, i.e.\ from any rule state for a rule $r$ with a slot $i$ matching $h$.
To facilitate deletion, we cache this information when $h$ is added to the rule state.
When $h$ is deleted, we can then efficiently determine the relevant rule states and delete $h$ from the hypothesis index $H$ at slot $i$ and from the match index $M$ at all slots $j ≥ i$.

\para{Example 1.}
As a first example, consider a rule state for the transitivity rule $r₁ : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
$h₁ : b ≤ c$ and $h₂ : a ≤ b$.
The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.

@@

\para{Example 2.}
We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot~2.
There are no matches yet, so there is nothing to extend.
We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the following rule state:
\begin{align*}
  H_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}};
  \end{tikzpicture}
  &&
  H_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}};
  \end{tikzpicture}
  &&
  M_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$a$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$b$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align*}

@@

\para{Unification.}
In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
As a result, unifying premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
Our technique therefore seeks to minimise the number of unifications performed.

@@

\para{RPINF.}
Since the hypothesis index $H$ and the match index $M$ of a rule state have the same domain, we combine them into one map.
This map, whose domain is a set of triples $(x, i, t)$, is represented by three nested persistent hash maps.

@@

\para{Match Equivalence.}
Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.

\para{Redundant Hypotheses.}
When we find a complete match, we add a new hypothesis of the corresponding type $T$ to the goal, but only if there is no other hypothesis of type $T$ already present.
To efficiently implement this check, we store in the forward match state a set containing the RPINFs of all hypotheses that have been added to (and not subsequently removed from) the state.
We modified the naive forward reasoning implementation to use the same mechanism.

\para{Lazy Insertion.}
Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
We therefore delay insertions until the index has selected at least one hypothesis for each slot.
This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process hypotheses for $a$ if there is, in fact, an involutive function in the context.

@@

\para{Transitivity Benchmark.}
Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$, where either $a = 0$ (yielding easy unification problems) or $a = 100$.
Saturating this goal adds $n(n-1)/2$ hypotheses to the context.

@@

\para{Independence Benchmark.}
In the last example, the incremental algorithm performs well partly because each match at level~1 can be combined with many hypotheses at level~2 without processing the level~1 hypothesis repeatedly.
We now consider a benchmark with many independent rules where this kind of reuse does not occur.

@@

\para{Depth Benchmark.}
Our final benchmark concerns situations where many rules match almost, but not quite.
We use 100 copies of a rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
In other words, the hypotheses match all premises with compatible substitutions, except the $k^{\text{th}}$ premise from the back.
Since the naive algorithm processes the premises back-to-front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.

@@

\para{White-Box Tactics.}
Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{PVS} have special support for forward reasoning.
Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.

@@

\para{SATCHMO.}
The SATCHMO first-order theorem prover~\cite{SATCHMO} uses a search procedure that treats a given set of implications as forward rules and uses a case-splitting rule to perform case analysis on disjunctive hypotheses.
SATCHMO-style provers may benefit from our incremental approach if the set of implications is large.
However, since unification is comparatively cheap in first-order logic, it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the forward state.
SATCHMO also employs an optimisation that \enquote*{focuses} forward reasoning on the hypothesis established by the last case split, providing a different form of incrementality.

\para{Knowledge-Based Systems.}
Some knowledge-based reasoning systems, e.g.\ Algernon~\cite{Algernon} and EYE~\cite{EYE}, support combined forward and backward reasoning.
However, to our knowledge, no such system allows non-forward rules to change the assumptions available to forward rules, so they do not require incrementality.
Nevertheless, our technique bears some resemblance to the RETE algorithm~\cite{RETE} for forward reasoning in large knowledge bases.
